Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-24 manuscript.rmd
1. Edited study site rainfall description 2. Edited discussion

We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual rainfall of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain month^-1^). Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].

@@

To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen.2018]. In moist forests $\Psi$~leaf~ equilibriates with $\Psi$~soil~ at night and declines through the day as plants transpire. $\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012]. We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\Psi$~soil~ as a proxy for the soil water limitation on growth. In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. This approach ignores the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause an assignment of a slightly deeper ERD to a species. Future work should therefore explore the influence of stem water storage on estimation of ERD (see Discussion). VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020]. The fraction of a tree's leaf cover present at a given time further limits it's growth potential. We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\
\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) to represent the impact of VPD on growth and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

@@

\
\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, (PLC)) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait. This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters). WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`). We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].

\
\   Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table. Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation. The model in @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the saprolite found in their dry forest study site (Jean Riotte, pers. comm.). The model of below-ground water-availability in this study (ELM-FATES) differs from the conceptual water balance model used by CT2018 and was calibrated with data on long-term soil moisture dynamics at a point location (0-0.15m) and plot-wide measurements by depth (0.15, 0.4 and 1 m), as well as, evapotranspiration and stream discharge. The model in CT2018 was calibrated only on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot. As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.

@@

The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests, which intrinsically rely on the parameterization of contrasting life history strategies [@Schieter:2012] and the simulation of competition between those strategies. We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). This offers the promise of a greater ability to parameterie the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics. The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

\
\   Our model estimated ERD as a single value, unlike the temporally dynamic ERD modeled by @Chitra-Tarak.2018, but future model developments should investigate seasonal dynamics of ERD, using corresponding calibration data on isotopically determined water-sourcing depths, as species are expected to shift to deeper soil water sources as shallow soils dry out [@Meinzer:1999]. Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes. Future studies could better constrain the ERD model with higher resolution growth data such as those from dendrometer bands. This is especially important for modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to hydrostatic stem flexing (xylem, phloem and bark hydration), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data and evaluation data ($\mathrm{\delta}$^2^H, Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and assumption of leaf presence during the wet season, and omits inter-annual variation. Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich tropical forests [@Park.2019]. In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curve ($\Psi$~soil~ vs. VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model. We recommend wide-spread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) [@Christoffersen:2016kd2d; @Sperry.1998] would be a step change from the current statistical models of ERD (this study and @Chitra-Tarak.2018). Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g. retention curves for roots), which limits immediate community-wide application, such models hold greater promise to improve our understanding of plant physiology [e.g. see @Johnson:2018fad].

>>>>

We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and data on stable water isotopes on tree species and soil from an independent study of tree water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 to create Gatun Lake and complete the Panama Canal. The entire island is forested and classified as a tropical moist forest in the Holdridge Life Zone system. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014]. Rainfall at BCI is seasonal with a mean annual total of `r round(rain.man.stats$rain)` mm ($\pm$ `r round(rain.man.stats$rain.sd)` SD; `r min(clim$Year)`-`r max(clim$Year)`) and a pronounced dry season from mid-December through April with a mean total of 292 mm [@Paton.2020]. Among the tree *species* found in the forest canopy, 28% are identified to be at least partly deciduous during the dry season, but this amounts to a maximum of 6.3% *trees* in the canopy that are deciduous, and the percentage of total forest crown area that is deciduous is estimated at 9.7% [@Condit:2000].

@@

To estimate effective rooting depth, we tested alternate models for growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. There are many intrinsic and extrinsic factors that limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Rhizosphere soil water potential ($\Psi$~soil~) down-regulates plant tissue water potentials [@Christoffersen:2016kd2d]. In moist forests $\Psi$~leaf~ equilibrates with $\Psi$~soil~ at night and declines through the day as plants transpire. $\Psi$~leaf~ in turn controls leaf conductance ($K_\textrm{leaf}$), which in turn controls stomatal conductance and photosynthetic rates [@Sack.2012]. We therefore used maximum diurnal leaf conductance ($K_\textrm{leaf}$) as determined by $\Psi$~soil~ as a proxy for the soil water limitation on growth. In our ERD models $K_\textrm{leaf}$ therefore incorporates the effects of rooting depth and rhizosphere water potential. This approach does not specifically incorporate the buffering effect of stem water storage on $K_\textrm{leaf}$, which may cause the model to slightly overestimate ERD (see Discussion). VPD has a positive impact on transpiration and thus photosynthesis and growth upto a VPD threshold, after which VPD has a negative impact [@Grossiord.2020]. The fraction of a tree's leaf cover present at a given time further limits it's growth potential. We therefore regressed species-specific growth in a census interval against census interval means of multiplicative or additive effects of one or more factors calculated daily: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. At our study site, due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\
\    To account for the non-linear impact of VPD on growth, we used a locally derived polynomial relationship between gross primary productivity (GPP) and VPD (see Model parametarisation) and used GPP standardised between 0 and 1 (hereafter, VPD*) in Eq. \@ref(eq:M-chosen).

@@

\
\   We found a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA and WSG) and parameters of leaf vulnerability (or percent loss of conductance, PLC) curves for a few species ($n$ = `r length(sp.exp.param$sp)`), a 'hard' trait. This allowed us to scale-up the leaf vulnerability curves at the community scale ($n$ = `r length(ab.table$Species)` species, see Table S\@ref(tab:ab-param) for model predicted parameters). WSG alone explained `r round(b.wsg.r2*100,0)`% of the variance in $B$ ($p$ `r  b.wsg.p`). We find that leaf conductance declined more rapidly with declining $\Psi$~leaf~ for species with higher values of WSG and LMA (Fig. S\@ref(fig:obs-plc-curves)). Whether such scaling relationships between hard and soft traits hold true for other forests, and their wider applicability, remains to be seen as data on PLC curves are limited in tropical forests, but supportive and counter examples both exist [@McCulloh.2011; @Hoffmann.2011].

\
\   Another important piece of information for understanding effective rooting depth is stable water isotopes of xylem water and below-ground sources, in the vadose zone and water table. Although the interpretation of these data is still under research [@Deurwaerder.2020], they are critical for model calibration or validation. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers and the spatial variability within the plot. As in @Chitra-Tarak.2018, we also interpret our ERD estimates as revealing relative differences among species' effective rooting depths rather than in absolute terms.

@@

The relationships that we identify between above- and below-ground traits, the water environment vertically below-ground and mortality rates are important for representing diversity in Dynamic Global Vegetation Models, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies. We found that LMA and WSG were strong predictors of PLC curves for leaves, but not for stems (not shown). This offers the promise of a greater ability to parameterise the 'hard' hydraulic traits with the abundant soft trait data, thus allowing much better representation of forest hydrodynamics. The relationships between ERD and above-ground hydraulic traits that we find will provide an important insight on how to model rooting depths and their coordination or trade-offs with other traits, including hydraulic vulnerability, in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

\
\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decouples climate events and demographic outcomes. Future studies could better constrain the ERD model with higher resolution growth data such as those from dendrometer bands. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also changes due to reversible stem dehydration (xylem, phloem and bark), the importance of which increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2015]. Use of dendrometer band data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018], but may also be an avenue to account for stem water storage and temporal dynamics of ERD, a limitation in our current approach. Three of our exploratory ERD models included seasonality in leaf area fraction, but we did not select them as they worsened the fit with growth data for most species (Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may be adequate to explain growth, but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf fall data and the backtracked timinng of leaf-gain from leaf life-time, and omits inter-annual variation. Future studies could further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019]. In forests such as BCI where understory tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees of the same size class, an issue we minimized by limiting ERD estimates to canopy trees. In sites where water does not limit growth, it may not be possible to parametarize ERD models from growth dynamics, and other direct avenues such as sap-flux dynamics could be explored. To estimate topographic variation in ERD within or across species, future studies may use soil moisture dynamics from a distributed hydrological model [e.g. @Schwantes:2018]. We note that data availability on soil water retention curve ($\Psi$~soil~ vs. VWC), hydraulic conductivity (K~soil~) by depth, soil moisture by depth, stream discharge and evapotranspiration were important for effective calibration of our 1-D hydrological model. We recommend wide-spread and coordinated collection of these variables as well as water-table levels in forest-inventory sites to allow for estimation of trees' water environments. While our model empirically predicts growth via $\Psi$~leaf~ from $\Psi$~soil~ at a specific depth, models that mechanistically account for $\Psi$~soil~ and $\Psi$~root~ over the entire rooting profile and then $\Psi$~stem~ and $\Psi$~leaf~, accounting for water content and embolised tissues, and downstream photosynthetic pathways and non-water limiting processes are likely to be more accurate in predicting growth [@Christoffersen:2016kd2d; @Sperry.1998]. Although data needs for parameterizing such models [@Christoffersen:2016kd2d; @Sperry.1998] could be greater (e.g. retention curves, capacitance for roots, stems and leaves), which increases parameter uncertainty in community-wide application, such models hold greater promise in improving our understanding of plant physiology [e.g. see @Johnson:2018fad].
